\textbf{Socio-demographic determinants of COVID-19 vaccine uptake in Ontario: Exploring differences across the Health Region model}

### 2023-07-28 manuscript_v7/main.tex
updated conclusion and discussion

\date{}
\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, interior hidden]}{\end{tcolorbox}}\fi

@@

Reasons given for this inequality in vaccination uptake have included
medical mistrust due to systemic medical
racism\textsuperscript{\protect\hyperlink{ref-stoler2021}{17},\protect\hyperlink{ref-mosby2021}{22}},
mistrust in
vaccines\textsuperscript{\protect\hyperlink{ref-willis2021}{15}}, and
the existence of health misinformation and
disinformation\textsuperscript{\protect\hyperlink{ref-mosby2021}{22}--\protect\hyperlink{ref-freeman2020}{24}}.
However, it is important to consider that vaccination uptake can also
vary geographically across the administrative or political subdivisions
within a country. For example, differences in COVID-19 vaccination rates
have been associated with differences in attitudes towards vaccination
in administrative areas in New York and
Chicago\textsuperscript{\protect\hyperlink{ref-malik2020}{25}}, scarcity
of vaccination facilities in areas where underrepresented groups reside
in Toronto\textsuperscript{\protect\hyperlink{ref-bogoch2022}{7}}, as
well as an absence of prioritization of areas inhabited by vulnerable
groups in the southeastern region of the
US\textsuperscript{\protect\hyperlink{ref-nguyen2021}{26}}. Other
studies have also shown heterogeneity in vaccine uptake within counties
in the
US\textsuperscript{\protect\hyperlink{ref-mollalo2021}{27}--\protect\hyperlink{ref-bhuiyan2022}{30}},
indicating that accounting for geographical differences in vaccination
can help predict patterns of booster uptake in Scottish
communities\textsuperscript{\protect\hyperlink{ref-wood2022}{31}}.
Therefore, the study of vaccination uptake across administrative of
political areas can be useful to highlight differences that governments
or public health agencies might need to address. However, in the case of
Canada, there is a limited amount of studies that have analyzed
geographical differences in vaccination. Existing studies in this area
have focused on differences within certain cities, such as
Toronto\textsuperscript{\protect\hyperlink{ref-choi2021}{32}}, or
Montréal\textsuperscript{\protect\hyperlink{ref-mckinnon2021}{33}}), or
have explored differences between
provinces-at-large\textsuperscript{\protect\hyperlink{ref-guay2022}{19}},
but to our knowledge, there are no studies that analyze differences in
vaccination uptake at the intra-provincial level. Such need is specially
important in the context of Canada's pandemic response goals, which have
been to minimize serious illness and deaths while minimizing societal
disruption\textsuperscript{\protect\hyperlink{ref-naci2023}{34}}.
Analyzing differences in vaccination uptake within the provinces can aid
to identify inequalities that might exist and that need to be addressed
before the advent of another pandemic.

@@

Furthermore, there is an ongoing challenge for the health system of the
province with regard to personalized healthcare for marginalized
individuals. For example, the West Health Region has only two Aboriginal
Health Access Centres (community-led primary healthcare organizations
focused on First Nations, Métis, and Inuit communities) to provide care
to an estimated population of 100,000 Indigenous individuals living in
the area\textsuperscript{\protect\hyperlink{ref-ontariohealth}{73}}.
Lack of access to personalized healthcare affects individuals that may
mistrust the traditional healthcare system due to systemic racism or
oppression, which is known to be the case for Indigenous and Black
individuals in Canada, as these rationales have been associated with
observed lower vaccination rates among these
groups\textsuperscript{\protect\hyperlink{ref-smylie2022}{74},\protect\hyperlink{ref-eissa2021}{75}}.
Taken together, this suggests that healthcare disparities specific to
these underrepresented groups in certain parts of the province impacted
vaccination uptake, and highlights the need of investments in the Health
Regions focused on resources, infrastructure, and specially personnel
that can deliver personalized care to marginalized communities, as it
has been shown that such efforts have improved trust in vaccination in
underrepresented groups
elsewhere\textsuperscript{\protect\hyperlink{ref-schafferderoo2020}{76}}.

@@

Additionally, our analysis did not consider the North West and North
East Health Regions, due to the low number of entries from these areas
in the survey (Figure~\ref{fig-map}). Low representation is expected as
these regions as they only account for 5\% of the total population of
Ontario. However, these areas have the highest proportion of Indigenous
inhabitants\textsuperscript{\protect\hyperlink{ref-ontariohealth}{73}}.
In the context of personalized care, there is a need to collect data
that focuses on these Health Regions where additional health disparities
might be present and possibly understudied.

The results in this study are based on self-reported data, where bias
might be present. However, in the context of COVID-19, it has been shown
that good agreement exists between self-reported and documented
vaccination
status\textsuperscript{\protect\hyperlink{ref-stephenson2022}{77}}, we
believe that our data was able to provide a valid sample of vaccination
uptake in the province. This is supported by the statistically
significant higher vaccination odds that were identified for January of
2022 in the model, which are consistent with province-wide trends
reported by Public Health Ontario (which show a 4\% increase between
early December and January, in contrast to a 2.5\% increase between
October and
November\textsuperscript{\protect\hyperlink{ref-ontario-covid}{78}});
however, the short time window constitutes essentially a ``snapshot''
view of the evolution of the disease, and additional data would be
needed to obtain estimates per racial/ethnic group over time across all
Health Regions that can help inform the existence of other health
disparities.

@@

The evidence collected in this study shows that differences in the
uptake of the first dose of the COVID-19 vaccine by underrepresented
individuals in the province of Ontario were the result of complex
relationships between infrastructure, resources, occupations, geography,
and socio-economic factors. Specifically, we identified that between the
last trimester of 2021 and January of 2022 individuals that identified
as Arab or Middle Eastern, Black, Latin American, of a mixed background,
or that belonged to other racial groups had significantly lower odds of
vaccination when compared to White/Caucasian individuals in Ontario.

However, although overall vaccination uptake was lower in these
underrepresented individuals in the province as a whole, the odds of
vaccination differed within certain income levels and Health Regions in
these groups, specially for those that reported a low household income
and that had higher vaccination odds that individuals in the same
racial/ethnic group with a higher income. These results highlight the
complex landscape of the province, where varying degrees of rurality
exist in conjunction with a socio-demographic makeup that is unique to
each of the Health Regions.

Such complexity presents an ongoing challenge for the healthcare system
of Ontario, and in particular, to the Health Region model which will
need to adapt and build on the LHIN model with the aim of providing an
integrative approach that can serve those that are most vulnerable
across all the Health Regions of the province. In particular,
personalized care is an area that could be further developed to improve
vaccination uptake in underrepresented individuals in the future. As we
have shown above, there are currently a limited number of centers
focused on community-led healthcare in the province, and our results
showed the existence of differences in vaccination uptake between
certain Health Regions. Improving personalized care for underrepresented
individuals then, can serve as a point to improve trust and facilitate
vaccine access in these marginalized communities.

Furthermore, in the event of a future pandemic or public health
emergency in the province in the years ahead, PHO and OH will need to
collaborate on the implementation of public health measures (such as
vaccination campaigns) building on the experience gained during the
COVID-19 vaccination rollout. So far, the COVID-19 pandemic served as a
stress test for both agencies, showing the necessity from decision
makers to clearly define how PHO and OH will interact in the future in
order to ensure that Ontarians can benefit by virtue of a robust
collaboration between both agencies.

Finally, the recent nature in the adaption of the Health Region poses a
challenge for researchers in the acquisition of data and information
that can be used to analyze the performance of the new model. From one
side, the Health Regions have not been incorporated as part of Census
data (LHINs were considered before in the Census), and this impacts the
amount of data that can be obtained. On the other hand, evaluations of
the Health Region model are at the moment not considered in the Annual
Reports of the Auditor General of Ontario, which was critical in the
past to identify limitations in the LHIN model. Currently, the only
demographic information available for each Health Region is provided by
Ontario Health (the agency that administers the Health Regions) but this
information only provides general estimates that do not allow for
detailed analyses of performance indicators (such as hospitalizations,
readmissions, and trends in chronic disease incidence) between the
Regions. Therefore, there is a pressing need for open data that can be
used by researchers and decision-makers to assess the performance of the
Health Region model.

@@

\leavevmode\vadjust pre{\hypertarget{ref-ontariohealth}{}}%
\CSLLeftMargin{73. }%
\CSLRightInline{\emph{{A}nnual {B}usiness {P}lan 2022/23}. Ontario
Health;
\url{https://www.ontariohealth.ca/sites/ontariohealth/files/2022-05/OHBusinessPlan22_23.pdf};
2022.}

\leavevmode\vadjust pre{\hypertarget{ref-smylie2022}{}}%
\CSLLeftMargin{74. }%
\CSLRightInline{Smylie J, McConkey S, Rachlis B, et al. Uncovering
{SARS}-{COV}-2 vaccine uptake and {COVID}-19 impacts among first
nations, inuit and m{é}tis peoples living in toronto and london,
ontario. \emph{Canadian Medical Association Journal}.
2022;194(29):E1018-E1026.
doi:\href{https://doi.org/10.1503/cmaj.212147}{10.1503/cmaj.212147}}

\leavevmode\vadjust pre{\hypertarget{ref-eissa2021}{}}%
\CSLLeftMargin{75. }%
\CSLRightInline{Eissa A, Lofters A, Akor N, Prescod C, Nnorom O.
Increasing {SARS}-{CoV}-2 vaccination rates among black people in
canada. \emph{Canadian Medical Association Journal}.
2021;193(31):E1220-E1221.
doi:\href{https://doi.org/10.1503/cmaj.210949}{10.1503/cmaj.210949}}

\leavevmode\vadjust pre{\hypertarget{ref-schafferderoo2020}{}}%
\CSLLeftMargin{76. }%
\CSLRightInline{DeRoo SS, Pudalov NJ, Fu LY. Planning for a {COVID}-19
vaccination program. \emph{{JAMA}}. 2020;323(24):2458.
doi:\href{https://doi.org/10.1001/jama.2020.8711}{10.1001/jama.2020.8711}}

\leavevmode\vadjust pre{\hypertarget{ref-stephenson2022}{}}%
\CSLLeftMargin{77. }%
\CSLRightInline{Stephenson M, Olson SM, Self WH, et al. Ascertainment of
vaccination status by self-report versus source documentation: Impact on
measuring {COVID}-19 vaccine effectiveness. \emph{Influenza and Other
Respiratory Viruses}. 2022;16(6):1101-1111.
doi:\href{https://doi.org/10.1111/irv.13023}{10.1111/irv.13023}}

\leavevmode\vadjust pre{\hypertarget{ref-ontario-covid}{}}%
\CSLLeftMargin{78. }%
\CSLRightInline{{Ontario COVID-19 Data Tool}. Accessed February 27,
2023.
\url{https://www.publichealthontario.ca/en/data-and-analysis/infectious-disease/covid-19-data-surveillance/covid-19-data-tool?tab=vaccine}}

>>>>

\date{}
\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, frame hidden, enhanced, breakable, sharp corners]}{\end{tcolorbox}}\fi

@@

Reasons given for this inequality in vaccination uptake have included
medical mistrust due to systemic medical
racism\textsuperscript{\protect\hyperlink{ref-stoler2021}{17},\protect\hyperlink{ref-mosby2021}{22}},
mistrust in
vaccines\textsuperscript{\protect\hyperlink{ref-willis2021}{15}}, and
the existence of health misinformation and
disinformation\textsuperscript{\protect\hyperlink{ref-mosby2021}{22}--\protect\hyperlink{ref-freeman2020}{24}}.
However, it is important to consider that vaccination uptake can also
vary geographically across the administrative or political subdivisions
within a country. For example, differences in COVID-19 vaccination rates
have been associated with differences in attitudes towards vaccination
in administrative areas in New York and
Chicago\textsuperscript{\protect\hyperlink{ref-malik2020}{25}}, scarcity
of vaccination facilities in areas where underrepresented groups reside
in Toronto when the provincial vaccination program
started\textsuperscript{\protect\hyperlink{ref-bogoch2022}{7}}, as well
as an absence of prioritization of areas inhabited by vulnerable groups
in the southeastern region of the
US\textsuperscript{\protect\hyperlink{ref-nguyen2021}{26}}. Other
studies have also shown heterogeneity in vaccine uptake within counties
in the
US\textsuperscript{\protect\hyperlink{ref-mollalo2021}{27}--\protect\hyperlink{ref-bhuiyan2022}{30}},
indicating that accounting for geographical differences in vaccination
can help predict patterns of booster uptake in Scottish
communities\textsuperscript{\protect\hyperlink{ref-wood2022}{31}}.
Therefore, the study of vaccination uptake across administrative of
political areas can be useful to highlight differences that governments
or public health agencies might need to address. However, in the case of
Canada, there is a limited amount of studies that have analyzed
geographical differences in vaccination. Existing studies in this area
have focused on differences within certain cities, such as
Toronto\textsuperscript{\protect\hyperlink{ref-choi2021}{32}}, or
Montréal\textsuperscript{\protect\hyperlink{ref-mckinnon2021}{33}}), or
have explored differences between
provinces-at-large\textsuperscript{\protect\hyperlink{ref-guay2022}{19}},
but to our knowledge, there are no studies that analyze differences in
vaccination uptake at the intra-provincial level. Such need is specially
important in the context of Canada's pandemic response goals, which have
been to minimize serious illness and deaths while minimizing societal
disruption\textsuperscript{\protect\hyperlink{ref-naci2023}{34}}.
Analyzing differences in vaccination uptake within the provinces can aid
to identify inequalities that might exist and that need to be addressed
before the advent of another pandemic.

@@

Furthermore, there is an ongoing challenge for the health system of the
province with regard to personalized healthcare for marginalized
individuals. Indeed, previous studies have pointed out mistrust in the
traditional healthcare system (due to systemic racism or oppression) as
a rationale for lower vaccination rates in the case of Indigenous and
Black individuals in
Canada\textsuperscript{\protect\hyperlink{ref-smylie2022}{73},\protect\hyperlink{ref-eissa2021}{74}}.
Other studies have also shown a need for increased culturally responsive
care for other underrepresented groups such as Black women and and other
immigrant groups in Canada and
Ontario\textsuperscript{\protect\hyperlink{ref-etowa2021}{75},\protect\hyperlink{ref-patel2019}{76}}.
Considering that for example, the West Health Region has only two
Aboriginal Health Access Centres (community-led primary healthcare
organizations focused on First Nations, Métis, and Inuit communities) to
provide care to an estimated population of 100,000 Indigenous
individuals living in the
area\textsuperscript{\protect\hyperlink{ref-ontariohealth}{77}}, it is
possible that the limited personalized healthcare for underrepresented
groups in certain parts of the province impacted vaccination uptake as
well, and highlights the need of investments in the Health Regions
focused on resources, infrastructure, and specially personnel that can
deliver personalized care to marginalized communities, as it has been
shown that such efforts have improved trust in vaccination in
underrepresented groups
elsewhere\textsuperscript{\protect\hyperlink{ref-schafferderoo2020}{78}}.

@@

Additionally, our analysis did not consider the North West and North
East Health Regions, due to the low number of entries from these areas
in the survey (Figure~\ref{fig-map}). Low representation is expected as
these regions as they only account for 5\% of the total population of
Ontario. However, these areas have the highest proportion of Indigenous
inhabitants\textsuperscript{\protect\hyperlink{ref-ontariohealth}{77}}.
In the context of personalized care, there is a need to collect data
that focuses on these Health Regions where additional health disparities
might be present and possibly understudied.

The results in this study are based on self-reported data, where bias
might be present. However, in the context of COVID-19, it has been shown
that good agreement exists between self-reported and documented
vaccination
status\textsuperscript{\protect\hyperlink{ref-stephenson2022}{79}}, we
believe that our data was able to provide a valid sample of vaccination
uptake in the province. This is supported by the statistically
significant higher vaccination odds that were identified for January of
2022 in the model, which are consistent with province-wide trends
reported by Public Health Ontario (which show a 4\% increase between
early December and January, in contrast to a 2.5\% increase between
October and
November\textsuperscript{\protect\hyperlink{ref-ontario-covid}{80}});
however, the short time window constitutes essentially a ``snapshot''
view of the evolution of the disease, and additional data would be
needed to obtain estimates per racial/ethnic group over time across all
Health Regions that can help inform the existence of other health
disparities.

@@

The evidence collected in this study shows differences in the uptake of
the first dose of the COVID-19 vaccine in Ontario between October 2021
and January 2022 in underrepresented groups, which had significantly
lower odds of vaccination when compared to White/Caucasian individuals.
However, although overall vaccination uptake was lower in
underrepresented individuals in the province as a whole, the odds of
vaccination differed within certain income levels and Health Regions in
these groups. Specifically, those that reported a low household income
had higher vaccination odds when compared to individuals in the same
racial/ethnic group in a higher income bracket. These results highlight
the complex landscape of the province, where varying degrees of rurality
exist in conjunction with a socio-demographic makeup that is unique to
each of the Health Regions.

Personalized care is an area that could be further developed to improve
vaccination uptake in underrepresented individuals in the future. As we
have shown in the discussion, there are currently a limited number of
centers focused on community-led healthcare in the province. Our results
showed the existence of differences in vaccination uptake between
certain Health Regions and therefore, improving personalized care for
underrepresented individuals can serve as a point to improve trust and
facilitate vaccine access in these marginalized communities.

Overall, there are ongoing challenges for healthcare access across
different segments of the population of Ontario. Future studies can be
focused at more in-depth analyses of vaccination uptake within the
different socio-demographic groups in each Health Region, in order to
provide decision-makers with information that can serve to carefully
consider and address how to improve vaccine access to marginalized
communities in Ontario.

@@

\leavevmode\vadjust pre{\hypertarget{ref-smylie2022}{}}%
\CSLLeftMargin{73. }%
\CSLRightInline{Smylie J, McConkey S, Rachlis B, et al. Uncovering
{SARS}-{COV}-2 vaccine uptake and {COVID}-19 impacts among first
nations, inuit and m{é}tis peoples living in toronto and london,
ontario. \emph{Canadian Medical Association Journal}.
2022;194(29):E1018-E1026.
doi:\href{https://doi.org/10.1503/cmaj.212147}{10.1503/cmaj.212147}}

\leavevmode\vadjust pre{\hypertarget{ref-eissa2021}{}}%
\CSLLeftMargin{74. }%
\CSLRightInline{Eissa A, Lofters A, Akor N, Prescod C, Nnorom O.
Increasing {SARS}-{CoV}-2 vaccination rates among black people in
canada. \emph{Canadian Medical Association Journal}.
2021;193(31):E1220-E1221.
doi:\href{https://doi.org/10.1503/cmaj.210949}{10.1503/cmaj.210949}}

\leavevmode\vadjust pre{\hypertarget{ref-etowa2021}{}}%
\CSLLeftMargin{75. }%
\CSLRightInline{Etowa J, Hyman I. Unpacking the health and social
consequences of {COVID}-19 through a race, migration and gender lens.
\emph{Canadian Journal of Public Health}. 2021;112(1):8-11.
doi:\href{https://doi.org/10.17269/s41997-020-00456-6}{10.17269/s41997-020-00456-6}}

\leavevmode\vadjust pre{\hypertarget{ref-patel2019}{}}%
\CSLLeftMargin{76. }%
\CSLRightInline{Patel A, Dean J, Edge S, Wilson K, Ghassemi E. Double
burden of rural migration in canada? Considering the social determinants
of health related to immigrant settlement outside the cosmopolis.
\emph{International Journal of Environmental Research and Public
Health}. 2019;16(5):678.
doi:\href{https://doi.org/10.3390/ijerph16050678}{10.3390/ijerph16050678}}

\leavevmode\vadjust pre{\hypertarget{ref-ontariohealth}{}}%
\CSLLeftMargin{77. }%
\CSLRightInline{\emph{{A}nnual {B}usiness {P}lan 2022/23}. Ontario
Health;
\url{https://www.ontariohealth.ca/sites/ontariohealth/files/2022-05/OHBusinessPlan22_23.pdf};
2022.}

\leavevmode\vadjust pre{\hypertarget{ref-schafferderoo2020}{}}%
\CSLLeftMargin{78. }%
\CSLRightInline{DeRoo SS, Pudalov NJ, Fu LY. Planning for a {COVID}-19
vaccination program. \emph{{JAMA}}. 2020;323(24):2458.
doi:\href{https://doi.org/10.1001/jama.2020.8711}{10.1001/jama.2020.8711}}

\leavevmode\vadjust pre{\hypertarget{ref-stephenson2022}{}}%
\CSLLeftMargin{79. }%
\CSLRightInline{Stephenson M, Olson SM, Self WH, et al. Ascertainment of
vaccination status by self-report versus source documentation: Impact on
measuring {COVID}-19 vaccine effectiveness. \emph{Influenza and Other
Respiratory Viruses}. 2022;16(6):1101-1111.
doi:\href{https://doi.org/10.1111/irv.13023}{10.1111/irv.13023}}

\leavevmode\vadjust pre{\hypertarget{ref-ontario-covid}{}}%
\CSLLeftMargin{80. }%
\CSLRightInline{{Ontario COVID-19 Data Tool}. Accessed February 27,
2023.
\url{https://www.publichealthontario.ca/en/data-and-analysis/infectious-disease/covid-19-data-surveillance/covid-19-data-tool?tab=vaccine}}
